Separating Use and Reuse to Improve Both

### 2017-05-08 FACS/Intro.tex
Intro

Historically, there has been a lot of focus on the importance of
separing subtyping from subclassing~\cite{cook}.  This is claimed to be
good for code reuse, design and reasoning. 
While there is no problem
in subtyping without subclassing, in most OO languages code reuse
(inheritance/extends) implies subtyping in a fundamental way because 
of what we call the self-reference \emph{leaking problem}. This 
problem is illustrated by the following (Java) code:

@@

\paragraph{Classes in \name: A mechanism for code Use}
\name has a substantially different design from Java-like languages.
The concept of a class in \name provides a mechanism for code use
only. This means that in \name there is actually no subclassing, and
and classes are roughly equivalent to final classes in Java.  Thus,
compared to Java-like languages, the most noticeable difference is the
absence of the \Q@extends@ keyword in \name.

@@

\noindent Classes in \name use a different declaration style compared
to Java. Class declarations have a name (which must always start with
an uppercase letter) and a code literal, which is used to specify the
definitions of the class. For instance, in the class declaration
for \lstinline{A}, the name of the class is \lstinline{A} and the code 
literal associatted with the class (\Q@{ method int ma()   Utils.m(this) }@) contains the definitions associated to the
class. There are important differences to Java-like languages in the way 
classes and code-literals are type-checked, as we shall see next. 
However for this example, things still work in a similar way to Java.

The \name code above is fine, but there is no way to add a \Q@B@ reusing code of \Q@A@, since
\Q@A@ is designed for code \emph{use} and not \emph{reuse}. So, unlike
the Java code, there is no way to introduce a subclass
\lstinline{B}. At first, this may seem like a severe restriction, but
\name has a different mechanism for code \emph{reuse} that 
is more appropriate when code reuse is intended.

@@

\noindent Here \lstinline{ta} is a trait intended to replace the
original class \lstinline{A} so that the code of the method
\lstinline{ma} can be reused. Then the class \lstinline{A} 
is created by inheriting the code from the trait using the keyword 
{\bf Use}. Note that \use\ cannot contain class names: only trait
names are allowed.
That is, using a trait is the only way to induce code reuse.
Unfortunatelly, this code does not work, 
because \lstinline{Utils.m} requires an \lstinline{A} and \lstinline{this} in
\Q@ta@ has no knowlegde of \lstinline{A}.
%: in \name all declarations 
%are sequentially type-checked and if a definition occurs afterwards
%another, then it is not visible on all previous definitions.

@@

\item The second reason is that {\bf the type of self-references is the structural type of the
    code literal.} Unlike Java-like languages, where the body of a
  class declaration is intrinsically coupled with the class itself, in
  \name code literals are first-class\bruno{Is it appropriate to say
    that they are first class?} and are type-checked independently
  using a form of structural typing.\bruno{Say how we recover nominality?} 
\end{itemize}

@@

This code works: \Q@Utils@ is using an interface \Q@IA@ and the trait \Q@ta@
is implementing it. It is also possible to add a \Q@B@ as follows
\begin{lstlisting}
  B:Use ta, { method int mb(){return this.ma();} }
\end{lstlisting}
This also works.  \Q@B@ reuses the code of \Q@ta@, but has no knowledge of \Q@A@.
Since \Q@B@ reuses \Q@ta@, and \Q@ta@ implements \Q@IA@, also \Q@B@ implements \Q@IA@.

\paragraph*{Semantic of Use}
Abeit alternative semantic models for traits~\cite{} have been proposed,
here we use the flattening model.
This means that 
\begin{lstlisting}
A:Use ta
B:Use ta, { method int mb(){return this.ma();} }
\end{lstlisting}

@@

\subsection{Recursive Types}

>>>>

Historically, there has been a lot of focus on the importance of
separing subtyping from subclassing~\cite{cook}.  This is claimed to be
good for code reuse, design and reasoning. There are at
least two distinct situations where the separation of subtyping and 
subclassing is helpful.

\item {\bf Allowing reuse even when subtyping is impossible:} 
In some situations a subclass contains methods whose signatures 
are incompatible with the superclass, yet inheritance is still
possible. A typical example, which was illustrated by Cook et al., are 
classes with \emph{binary methods}~\cite{}.

\item {\bf Preventing unintended subtyping:} For certain classes we
  would like to inherit code without creating a subtype even if, from
  the typing point of view, subtyping is still possible. A typical
  example~\cite{} of this are methods for collection classes such as Sets and
  Bags. Bag implementations can often inherit 
  from Set implementations, and the interfaces of the two collection types are
  similar and type compatible. 
  However, from the logical point-of-view a Bag is \emph{not a
    subtype} of a Set.

\end{itemize}

Type systems based on structural typing can deal with the first
situation well, but not the second. Since structural subtyping
accounts for the types of the methods only, a Bag would be a subtype
of a Set if the two interfaces are type compatible. For dealing with
the second situation nominal subtyping is preferable. With nominal
subtyping an explicit subtyping relation must be signalled by the
programmer. Thus if subtyping is not desired, the idea is that 
programmer can simply not declare a subtyping relationship.

While there is no problem in subtyping without subclassing, in the design
of most nominal OO languages subtyping implies subtyping in a
fundamental way. This is because of what we call the self-reference
\emph{leaking problem}. The leaking problem is illustrated by the following
(Java) code:

@@

\paragraph{Classes in \name: A mechanism for code Use}
%\name has a substantially different design from Java-like languages.
The concept of a class in \name provides a mechanism for code use
only. This means that in \name there is actually no subclassing, and
and classes are roughly equivalent to final classes in Java.  Thus,
compared to Java-like languages, the most noticeable difference is the
absence of the \Q@extends@ keyword in \name.

@@

\noindent Classes in \name use a slightly different declaration style compared
to Java. In \name there is no \lstinline{class} keyword. 
Class declarations have a name (which must always start with
an uppercase letter) and a code literal, which is used to specify the
definitions of the class. For instance, in the class declaration
for \lstinline{A}, the name of the class is \lstinline{A} and the code 
literal associatted with the class (\Q@{ method int ma()   Utils.m(this) }@) contains the definitions associated to the
class. There are, however, some 
important differences to Java-like languages in the way 
classes and code-literals are type-checked, as we shall see next. 
Nevertheless, for this example, things still work in a similar way to Java.

The \name code above is fine, but there is no way to add a \Q@B@ reusing code of \Q@A@, since
\Q@A@ is designed for code \emph{use} and not \emph{reuse}. So, unlike
the Java code, introducing a subclass
\lstinline{B} is not possible. At first, this may seem like a severe restriction, but
\name has a different mechanism for code \emph{reuse} that 
is more appropriate when code reuse is intended.

@@

\noindent Here \lstinline{ta} is a trait intended to replace the
original class \lstinline{A} so that the code of the method
\lstinline{ma} can be reused. Then the class \lstinline{A} 
is created by inheriting the code from the trait using the keyword 
{\bf Use}. Note that \use\ cannot contain class names: only trait
names are allowed.
That is, using a trait is the only way to induce code reuse.
Unfortunatelly, this code does not work, 
because \lstinline{Utils.m} requires an \lstinline{A} and \lstinline{this} in
\Q@ta@ has no knowlegde of \lstinline{A}.
\bruno{Can we summarize in a table what are the differences between
  traits and classes? For example: 
traits: do not introduce types; units of reuse; cannot be instantiated
classes: introduce types; can be instantiated. 
Also it would be good to motivate those differences more from the 
technical point of view. For example: the fact that traits do not
introduce types is it mostly a philosophical difference, or does it 
actually buy you something from the technical aspect.
}
%: in \name all declarations 
%are sequentially type-checked and if a definition occurs afterwards
%another, then it is not visible on all previous definitions.

@@

\item The second reason is that {\bf the type of self-references is the self type of the
    code literal.} Unlike Java-like languages, where the body of a
  class declaration is intrinsically coupled with the class itself, in
  \name code literals are first-class\bruno{Is it appropriate to say
    that they are first class?} and are type-checked independently.
  \name has a notion of self-type, which is closely related to
  approaches such as ThisType~\cite{} and ...\cite{others}. Therefore
  in \name the self-reference does not have the type of the class
  being defined, but rather it has the self-type.
\end{itemize}

@@

This code works: \Q@Utils@ is using an interface \Q@IA@ and the trait \Q@ta@
is implementing it. Thus the self-type of the trait \Q@ta@ will be a
subtype of \Q@IA@. It is also possible to add a \Q@B@ as follows
\begin{lstlisting}
  B:Use ta, { method int mb(){return this.ma();} }
\end{lstlisting}
This also works.  \Q@B@ reuses the code of \Q@ta@, but has no knowledge of \Q@A@.
Since \Q@B@ reuses \Q@ta@, and \Q@ta@ implements \Q@IA@, also \Q@B@ implements \Q@IA@.

\paragraph*{Semantic of Use}
Albeit alternative semantic models for traits~\cite{} have been proposed,
here we use the flattening model.
This means that 
\begin{lstlisting}
A:Use ta
B:Use ta, { method int mb(){return this.ma();} }
\end{lstlisting}

@@

\subsection{Self Types, Binary Methods and the Need to Separate
  Inheritance from Subtyping}

So, far it appears that we have not gained much from decoupling
reuse from subtyping.

As discussed in Section~\ref{} the type of the self-reference is the
self-type. Like languages with ThisType, \name allows explicit
references to the self type. This is useful in several situations: 
For example it can be helpful in the presence of binary
methods~\cite{}, or methods that return an instance of the 
class being defined.
